---
title: "Solution to the exercise - from TF binding sites to position-specific scoring matrices"
author: "Jacques van Helden"
date: '`r Sys.Date()`'
output:
  html_document:
    fig_caption: yes
    highlight: zenburn
    theme: cerulean
    toc: yes
    toc_depth: 3
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  word_document: default
---


```{r knitr setup, include=FALSE,  eval=TRUE, echo=FALSE, warning=FALSE}
library(knitr)
knitr::opts_chunk$set(echo=TRUE, eval=TRUE, cache=FALSE, message=FALSE, warning=FALSE, comment = "")
```

## Introduction

In this document we show the solution of the exercise "From TF binding sites to position-specific weight matrices" using the R language. We display both the code and the results. This document was generated with R markdown [from_TFBS_to_PSSM_solutions.Rmd](from_TFBS_to_PSSM_solutions.Rmd).

## Site sequences

```{r sequences}
## Encode the sequences in a vector
sequences <- c( "AAAAACCG",
                "AAAACCGG",
                "AATTGGGG",
                "ATTGTGGG")
print(as.data.frame(sequences))
```



## Alignment matrix


```{r alignment matrix}
## Build an alignment matrix, with one row per site 
## and one column per aligned position
alignment <- t(as.data.frame(strsplit(sequences, "")))
colnames(alignment) <- 1:ncol(alignment)
row.names(alignment) <- paste(sep="_", "site", 1:nrow(alignment))

kable(alignment, align="c")

```

## 1. Count matrix

*Build a position-specific scoring matrix indicating the residue counts per position (count matrix).*

```{r count matrix computation}
## Compute a count matrix from the alignment table
counts <- data.frame()
residues <- c("A", "C", "G", "T")
for (r in residues) {
  counts <- rbind(counts, apply(alignment==r, 2, sum))
}
colnames(counts) <- 1:ncol(counts)
row.names(counts) <- residues

kable(counts, align="c")
```

## 2. Raw frequencies (pseudo-weight = 0)

*From this count matrix, derive a frequency matrix.*

```{r frequency matrix}
freq <- counts / apply(counts, 2, sum)
kable(freq, align="c")
```


## 3. Corrected frequencies

*Compute a pseudo-weight smoothed frequency matrix, with pseudo-weights of 1, 10 and 100, respectively.*


### Equal prior probabilities

In this exercise we use equiprobable nucleotides, prior residue frequencies are thus set to $p=0.25$ for each residue.

```{r equiprobable prior frequencies}

equal.priors <- vector()
equal.priors[residues] <- 1/length(residues)

kable(data.frame(equal.priors), align="c")
```


### Pseudo-weight = 1

```{r corrected frequencies with pseudo-weight 1}
k <- 1 ## Arbitrary value for the pseudo-weight

## Compute corrected frequencies
corrected.freq <- (counts + k/length(residues)) / apply(counts + k*0.25, 2, sum)

kable(corrected.freq, align="c", digits=3)
```
  
  
### Pseudo-weight = 10

```{r corrected frequencies with pseudo-weight 10}
k <- 10 ## Arbitrary value for the pseudo-weight

## Compute corrected frequencies
corrected.freq <- (counts + k/length(residues)) / apply(counts + k*0.25, 2, sum)

kable(corrected.freq, align="c", digits=3)
```
  
### Pseudo-weight = 100

```{r corrected frequencies with pseudo-weight 100}
k <- 100 ## Arbitrary value for the pseudo-weight

## Compute corrected frequencies
corrected.freq <- (counts + k/length(residues)) / apply(counts + k*0.25, 2, sum)

kable(corrected.freq, align="c", digits=3)
```
  
  
## 4. Weight matrices, no pseudo-weight, equal priors

*Compute the weight matrices corresponding to the frequency matrices with no pseudo-weight, assuming equiprobable residues.*

```{r weights with equal priors and no pseudo-weight}
## Compute weights
weights <- log(freq / 0.25)

kable(weights, align="c", digits=2)
```
  

## 5. Weight matrix with equal priors and pseudo-weight=1

```{r weights with equal priors}
k <- 1 ## Arbitrary value for the pseudo-weight

## Compute corrected frequencies
corrected.freq <- (counts + k/length(residues)) / apply(counts + k*0.25, 2, sum)
weights <- log(corrected.freq / 0.25)

kable(weights, align="c", digits=2)
```
  
## 6. Unequal priors

*Compute the weight matrices with a pseudo-weight=1 and the following prior probabilities: A=0.3, T=0.3, C=0.2, G=0.2.*


```{r define unequal priors}
## Define nucleotide-specific residue priors
unequal.priors <- c("A"=0.3, "C"=0.2, "G"=0.2, "T"=0.3)
kable(data.frame(unequal.priors))
```

```{r frequencies with unequal priors}

## Arbitrary value for the pseudo-weight
k <- 1 

## Compute corrected frequencies
corrected.freq <- (counts + k*unequal.priors) / apply(counts + k*unequal.priors, 2, sum)
kable(corrected.freq, align="c", digits=2, 
      caption=paste("Frequency matrix with unequal priors and pseudo-weight=",k))

```

```{r weights with unequal priors}
weights <- log(corrected.freq / unequal.priors)
kable(weights, align="c", digits=2, caption = paste("Weight matrix with unequal priors and pseudo-weight =",k))

```

## 7. Interpretation of the results.





